\documentclass{article}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathpazo}
\newcommand{\zeroty}{\mathsf{void}}
\newcommand{\abort}[1]{\mathsf{abort}(#1)}
\newcommand{\sumty}[2]{{#1}+{#2}}
\newcommand{\inleft}[1]{\mathsf{inl}(#1)}
\newcommand{\inright}[1]{\mathsf{inr}{(#1)}}
\newcommand{\sumcase}[5]{\mathsf{case}({#1};{#2}{.}{#3};{#4}{.}{#5})}
\newcommand{\idty}[3]{{#2}\mathbin{=_{#1}}{#3}}
\newcommand{\refl}[2]{\mathsf{refl}_{#1}({#2})}
\newcommand{\idelim}[4]{\mathsf{J}[{#1}]({#2};{#3}{.}{#4})}
\newcommand{\family}[2]{{#1}.{#2}}
\newcommand{\familytwo}[3]{\family{#1}{\family{#2}{#3}}}
\newcommand{\familythree}[4]{\family{#1}{\familytwo{#2}{#3}{#4}}}
\newcommand{\prodty}[3]{\prod_{{#1}{:}{#2}}{#3}}
\newcommand{\sigty}[3]{\sum_{{#1}{:}{#2}}{#3}}
\newcommand{\lam}[2]{\lambda {#1}{.}{#2}}
\newcommand{\app}[2]{{#1}({#2})}
\newcommand{\apppath}[1]{\mathsf{ap}_{#1}}
\newcommand{\univty}{\mathcal{U}}

\newcommand{\inverse}[1]{#1^{-1}}
\newcommand{\concat}[2]{{#1}\cdot{#2}}

\newcommand{\transport}[2]{\mathsf{trans}[#1](#2)}

\newtheorem{theorem}{Theorem}
\newtheorem{defn}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\begin{document}

\section{Co-product Paths}

As in the HoTT book, we wish to prove the following characterization of paths in a coproduct type $A+B$:
\begin{enumerate}
\item $\idty{\sumty{A}{B}}{\inleft{M}}{\inleft{M'}} \simeq \idty{A}{M}{M'}$;
\item $\idty{\sumty{A}{B}}{\inright{N}}{\inright{N'}} \simeq \idty{B}{N}{N'}$;
\item $\idty{\sumty{A}{B}}{\inleft{\_}}{\inright{\_}} \simeq \zeroty$;
\item $\idty{\sumty{A}{B}}{\inright{\_}}{\inleft{\_}} \simeq \zeroty$.
\end{enumerate}

Define the family $$u:\sumty{A}{B},v:\sumty{A}{B}\vdash F[u,v]:\univty$$ so that the following definitional equivalences hold:
\begin{enumerate}
\item $F[\inleft{M},\inleft{M'}]\equiv \idty{A}{M}{M'}$;
\item $F[\inright{M},\inright{M'}]\equiv \idty{A}{M}{M'}$;
\item $F[\inleft{\_},\inright{\_}]\equiv\zeroty$;
\item $F[\inright{\_},\inleft{\_}]\equiv\zeroty$.
\end{enumerate}
This is easily achieved by a nested case analysis on $u$ and $v$, respectively, with motive $\univty$ in each case.  The correspondence between the defnition of $F$ and the desired theorem is evident.  The point is that $F$ cancels the injections definitionally, so that no particular mention need be made of this in the proof.

\begin{lemma}
  \label{L-lemma}
  There is a term $L$ of type $\prodty{z}{\sumty{A}{B}}{F[z,z]}$ such that $\app{L}{\inleft{x}}\equiv \refl{A}{x}$ and $\app{L}{\inright{y}}\equiv \refl{B}{y}$.
\end{lemma}
\begin{proof}
  Construct $L$ by abstracting over $z$, then performing a case analysis on $z$:
  \begin{enumerate}
  \item $x:A\vdash \refl{A}{x}:F[\inleft{x},\inleft{x}]$.
  \item $y:B\vdash \refl{A}{y}:F[\inright{y},\inright{y}]$.
  \end{enumerate}
\end{proof}

We wish to show that $$\prodty{z,z'}{\sumty{A}{B}}{\idty{\sumty{A}{B}}{z}{z'}}\simeq F[z,z'],$$ from which the desired result follows immediately by definition of $F$.

First we exhibit the function
$$f:\prodty{z}{\sumty{A}{B}}{\prodty{z'}{\sumty{A}{B}}{\idty{\sumty{A}{B}}{z}{z'}\to F[z,z']}}$$
given as follows:
$$\lam{z}{\lam{z'}{\lam{p}{\idelim{\familythree{u}{v}{\_}{F[u,v]}}{p}{x}{\app{L}{x}}}}}.$$
Lemma~\ref{L-lemma} does all the work: if $M:\sumty{A}{B}$, then $$\app{\app{\app{f}{M}}{M}}{\refl{\sumty{A}{B}}{M}}\equiv \app{L}{M}.$$

We then exhibit a quasi-inverse for $f$,
$$g:\prodty{z}{\sumty{A}{B}}{\prodty{z'}{\sumty{A}{B}}{F[z,z']\to\idty{\sumty{A}{B}}{z}{z'}}}$$
given by a nested case analysis on $z$ and $z'$ based on the following data:
\begin{enumerate}
\item $x:A,x':A,z:F[\inleft{x},\inleft{x'}]\vdash \app{\apppath{\inleft{-}}}{z}:\idty{\sumty{A}{B}}{\inleft{x}}{\inleft{x'}}$;
\item $x:A,y':B,z:F[\inleft{x},\inright{y'}]\vdash \abort{z}:\idty{\sumty{A}{B}}{\inleft{x}}{\inright{y'}}$.
\item $y:B,y':B,z:F[\inright{x},\inright{x'}]\vdash \app{\apppath{\inright{-}}}{z}:\idty{\sumty{A}{B}}{\inright{x}}{\inright{x'}}$;
\item $x:A,y':B,z:F[\inright{x},\inleft{y'}]\vdash \abort{z}:\idty{\sumty{A}{B}}{\inright{x}}{\inleft{y'}}$.
\end{enumerate}
Here we are relying on the definitional properties of the family $F$ to justify the given typings.  Notice that the $\lambda$-abstraction of the third argument to $g$ must occur \emph{inside} the case analysis in order to propagate the correct branch information (\textit{cf}. the hacky treatment of this issue in Haskell's so-called GADT's.)

\begin{lemma} 
  \label{g-lemma}
  The following types are inhabited:
  \begin{enumerate}
  \item $x:A\vdash \_ : \idty{\sumty{A}{B}}{\app{\app{\app{g}{\inleft{x}}}{\inleft{x}}}{\refl{A}{x}}}{\refl{\sumty{A}{B}}{\inleft{x}}}$;
  \item $y:B\vdash \_ : \idty{\sumty{A}{B}}{\app{\app{\app{g}{\inright{x}}}{\inright{x}}}{\refl{B}{y}}}{\refl{\sumty{A}{B}}{\inright{y}}}$.
  \end{enumerate}
\end{lemma}

To complete the proof we need only exhibit witnesses to the fact that for $z,z':\sumty{A}{B}$, the function  $g'=\app{\app{g}{z}}{z'}$ is right and left inverse to the function $f'=\app{\app{f}{z}}{z'}$, up to higher homotopy.
\begin{enumerate}
\item $z:\sumty{A}{B},z':\sumty{A}{B},w:F(z,z')\vdash \alpha : \idty{F[z,z']}{\app{f'}{\app{g'}{w}}}{w}$.
\item $z:\sumty{A}{B},z':\sumty{A}{B},w:\idty{\sumty{A}{B}}{z}{z'}\vdash \beta : \idty{\idty{\sumty{A}{B}}{z}{z'}}{\app{g'}{\app{f'}{w}}}{w}$.
\end{enumerate}

The first is proved by a nested case analysis on $z$ and $z'$, using the definitional equivalences governing $F$, either aborting, or performing a path induction on $w$, appealing to Lemma~\ref{g-lemma} to complete the proof.  The second is proved by induction on $w$, using a case analysis on the generic $u:\sumty{A}{B}$ so that we may appeal to Lemma~\ref{g-lemma} to complete the proof.

\section{Application to Paths}

The idea is to adapt the above proof for coproducts, except using propositional equivalences for the family $F$ obtained from the assumption that $f$ has a quasi-inverse.

We are given $f:A\to B$ and the following data showing that $f$ has a quasi-inverse:
\begin{enumerate}
\item $\inverse{f}:B\to A$;
\item $\alpha:\prodty{a}{A}{\idty{A}{\app{\inverse{f}}{\app{f}{a}}}{a}}$;
\item $\beta:\prodty{b}{B}{\idty{B}{\app{f}{\app{\inverse{f}}{b}}}{b}}$.
\end{enumerate}

We are to show that $\apppath{f}$ has a quasi-inverse, which amounts to giving the following data:
\begin{enumerate}
\item $\inverse{\apppath{f}}$, which is taken to be $\lam{q}{\concat{\concat{\inverse{\app{\alpha}{a}}}{\app{\apppath{\inverse{f}}}{q}}}{\app{\alpha}{a'}}}$;
\item $\alpha':\prodty{a}{A}{\prodty{a'}{A}{\prodty{p}{\idty{A}{a}{a'}}{\idty{\idty{A}{a}{a'}}{\app{\inverse{\apppath{f}}}{\app{\apppath{f}}{p}}}{p}}}}$;
\item $\beta':\prodty{a}{A}{\prodty{a'}{A}{\prodty{q}{\idty{B}{\app{f}{a}}{\app{f}{a'}}}{\idty{\idty{B}{\app{f}{a}}{\app{f}{a'}}}{\app{\apppath{f}}{\app{\inverse{\apppath{f}}}{q}}}{q}}}}$.
\end{enumerate}

The construction of $\alpha'$ is given in the book.  It takes the form of a path induction on $p$, reducing the problem to the case of reflexivity for a generic $x$ of type $A$.  This presents no difficulties.

The construction of $\beta'$ is more difficult, because the conclusion involves $a$, $a'$, and $q$, whereas the type of $q$, on which one would consider to perform a path induction, is $\idty{B}{\app{f}{a}}{\app{f}{a'}}$.  This means that whatever motive, $F$, we choose for the path induction, the conclusion will be of the form $F[\app{f}{a},\app{f}{a'},q]$, and we must somehow ``cancel'' the $f$'s in order to get back $a$ and $a'$.  We can do this by exploiting the quasi-inverse we are given for $f$.

Thus define $u:B,v:B,w:\idty{B}{u}{v}\vdash F:\univty$ to be the following equality type:
\begin{displaymath}
  \idty{\idty{B}{u}{v}}{\app{\apppath{f}}{\concat{\concat{\inverse{\app{\alpha}{\app{\inverse{f}}{u}}}}{\app{\apppath{\inverse{f}}}{w}}}{\app{\alpha}{\app{\inverse{f}}{v}}}}}{w}
\end{displaymath}
If we use $F$ as the motive for a path induction over $q$ of type $\idty{B}{\app{f}{a}}{\app{f}{a'}}$, then we obtain an inhabitant of ${F[\app{f}{a},\app{f}{a'},q]}$, whichis just the equality type
\begin{displaymath}
  \idty{}
  {\app{\apppath{f}}{\concat{\concat{\inverse{\app{\alpha}{\app{\inverse{f}}{\app{f}{a}}}}}{\app{\apppath{\inverse{f}}}{q}}}{\app{\alpha}{\app{\inverse{f}}{\app{f}{a'}}}}}}
  {q}.
\end{displaymath}
Using data provided by the quasi-inverse for $f$, we may show that the left-hand side of this equation is
\begin{displaymath}
  \app{\apppath{f}}{\concat{\concat{\inverse{\app{\alpha}{a}}}{\app{\apppath{\inverse{f}}}{q}}}{\app{\alpha}{a'}}},
\end{displaymath}
which is the desired conclusion.

It remains to show, then, that 
\begin{displaymath}
  y:B \vdash \_ : F[y,y,\refl{B}{y}],
\end{displaymath}
which is to say that 
\begin{displaymath}
  y:B \vdash \_ : \idty{B}{\app{\apppath{f}}{\concat{\concat{\inverse{\app{\alpha}{y}}}{\app{\apppath{\inverse{f}}}{\refl{B}{y}}}}{\app{\alpha}{y}}}}{\refl{B}{y}}.
\end{displaymath}
Now $\app{\apppath{\inverse{f}}}{\refl{B}{y}}\equiv \refl{B}{\app{\inverse{f}}{y}}$, which is the unit of concatenation (up to higher homotopy), so it suffices to show
\begin{displaymath}
  y:B \vdash \_ : \idty{B}{\app{\apppath{f}}{\concat{\inverse{\app{\alpha}{y}}}{\app{\alpha}{y}}}}{\refl{B}{y}}.
\end{displaymath}
Applying the inverse law for paths, this is just
\begin{displaymath}
  y:B \vdash \_ : \idty{B}{\app{\apppath{f}}{\refl{B}{\app{\inverse{f}}{y}}}}{\refl{B}{y}}.
\end{displaymath}
Finally, $\app{\apppath{f}}{\refl{B}{\app{\inverse{f}}{y}}}\equiv \refl{B}{\app{f}{\app{\inverse{f}}{y}}}$, and we have
\begin{displaymath}
  y:B \vdash \_ : \idty{B}{\refl{B}{\app{f}{\app{\inverse{f}}{y}}}}{\refl{B}{y}},
\end{displaymath}
again using the quasi-inverse of $f$, which completes the argument.

\smallskip

Throughout I am tacitly using the principle that $$\transport{\family{x}{x}}{p}:A\to A'$$ whenever $p:\idty{\univty}{A}{A'}$, which is to say that one may transport objects of a type $A$ to an object of an equal type $A'$ in the same universe.  This move corresponds to the implicit uses of definitional equality of types in the characterization of the paths in a coproduct type.

\end{document}
